Equations & R Code

### 2016-11-02 manuscript/coexistence-stability_ms.Rmd
small changes to ms

#### Implementing the Storage Effect
For the storage effect to operate, we need species-specific responses to environmental variability, positive environment-competition (EC) covariance, and subadditive population growth (@Chesson1994; @Chesson2000a). 
EC covariance limits population growth of residents: in a favorable year, they are limited by strong intraspecific competition because their populations are large. 
However, a species at very low abundance can escape the limitations of EC covariance if it experiences favorable conditions when the residents it is competiting against experience (relatively) unfavorable conditions. 
Subadditive population growth buffers populations against large population decreases in unfavorable years.

>>>>

#### Implementing the Storage Effect
For the storage effect to operate, we need species-specific responses to environmental variability, positive environment-competition (EC) covariance, and subadditive population growth [@Chesson1994; @Chesson2000a]. 
EC covariance limits population growth of residents: in a favorable year, they are limited by strong intraspecific competition because their populations are large. 
However, a species at very low abundance can escape the limitations of EC covariance if it experiences favorable conditions when the residents it is competiting against experience (relatively) unfavorable conditions. 
Subadditive population growth buffers populations against large population decreases in unfavorable years.

### 2016-11-02 manuscript/coexistence-stability_ms.tex
small changes to ms

MacArthur (1955), Elton (1958), and even Darwin (Turnbull et al. 2013)
recognized that species can compensate for each other and stabilize
functioning in ecosystems subject to temporal variation in environmental
conditions. This idea underlies the ``insurance hypothesis'' (Yachi and
Loreau 1999), which suggests stability increases with diversity because
species respond dissimilarly to environmental conditions -- species A
has highest growth rates under conditions X whereas species B has
highest growth rates under conditions Y. More species confer temporal
stability by broadening the range of conditions under which the
community maintains function (Loreau 2010). Diverse models all predict a
positive relationship between species richness and ecosystem stability
(Lehman and Tilman 2000, Ives and Hughes 2002, Loreau and {{de
Mazancourt}} 2013), and experimental tests tend to support such a
prediction (Tilman et al. 2006, Hector et al. 2010).

However, the ability of biodiversity-ecosystem functioning (BEF)
experiments to represent real-world dynamics is the subject of debate
(Wardle 2016, Eisenhauer et al. 2016). Much of the debate centers around
the fact that species losses in BEF experiments are not allowed to be
offset by species gains. Theory typically suffers from the same
limitation, in that it often ignores the processes that could lead to
species gains in temporally variable environments (Loreau 2010, but see
Chesson et al. 2001).

Temporally fluctuating environmental conditions are an important
ingredient for stable species coexistence, both in theoretical models
(Chesson 2000, Chesson et al. 2004) and in natural communities
(C{á}ceres 1997, Descamps-Julien and Gonzalez 2005, Adler et al. 2006,
Angert et al. 2009). Such ``fluctuation-dependent'' coexistence (Chesson
2000) requires that species have unique environmental responses and that
environmental conditions vary enough for each coexisting species to
experience good and bad conditions. Thus, there is reason to expect
environmental variability to promote species richness when coexistence
is maintained by a fluctuation-dependent mechanism (Adler and Drake
2008). Of course, increasing environmental variability may also decrease
ecosystem stability through time by increasing the fluctuations of
individual species, regardless of species richness.

The countervailing effects of environmental variability present an
interesting paradox: increasing variability should decrease ecosystem
stability, but may also increase richness, which may offset the decrease
in stability. Such a paradox complicates predictions about how
ecosystems will respond as environmental conditions exceed historical
ranges of variability. The unknown net effect of environment variability
may be reflected in the mixed results from empirical studies on the
diversity-stability relationship. Observational tests of the
diversity-stability relationship, which require sampling across natural
diversity gradients, have yielded positive (Hautier et al. 2014),
neutral (Valone and Hoffman 2003, Cusson et al. 2015), and negative
(Sasaki and Lauenroth 2011) relationships. In a meta-analysis of
diversity-stability relationships, Jiang and Pu (2009) found no
significant evidence for an effect of species richness on ecosystem
stability from observational studies in terrestrial ecosystems. Thus,
there appears to be a gap between the consistency of theoretical studies
and the equivocation of empirical studies.

We argue this gap exists because the two bodies of theory that have
developed to explain species coexistence on the one hand, and
diversity-stability relationships on the other, have diverged. One
reason these two disciplines have diverged is because they have focused
on slightly different questions. Biodiversity-ecosystem stability
studies typically ask how ecosystem variability responds to different
levels of species richness at a given level of environmental variability
(reviewed in Kinzig et al. 2001, Loreau 2010), whereas coexistence
studies ask how the long term stability of species coexistence responds
to different levels of environmental variability (Chesson and Warner
1981).

To reconcile these two bodies of theory, we extend theory on the
relationship between species richness and ecosystem stability to cases
in which species coexistence explicitly depends on environmental
fluctuations and species-specific responses to environmental conditions.
We focus on the storage effect and relative nonlinearity using a general
consumer-resource model. We use the model to investigate two questions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Does the direction of the diversity-stability relationship remain
  positive when species coexistence is fluctuation-dependent? And,
\item
  When species coexistence is fluctuation-dependent, how does increasing
  environmental variability impact ecosystem stability?
\end{enumerate}

@@

To make this a storage effect model, we need to satisfy three
conditions: (1) the organisms must have a mechanism for persistence
under unfavorable conditions, (2) species must respond differently to
environmental conditions, and (3) the effects of competition on a
species must be more strongly negative in good years relative to
unfavorable years. Our model meets condition 1 because we include a
dormant stage with very low death rates. We satisfy condition 2 with our
model whenever \(\gamma\) is not perfectly correlated between species.
Lastly, our model meets condition 3 because condition 2 partitions
intraspecific and interspecific competition into different years. Thus,
during a high \(\gamma\) year for one species, resource uptake is also
inherently high for that species, which increases intraspecific
competition relative to interspecific competition. So, given adequate
variability in \(\gamma\), the inferior competitor can persist. We
created competitive hierarchies in the storage effect version of the
model by altering species' biomass conversion efficiencies
(\(\epsilon\))

@@

@@

To understand how fluctuation-dependent coexist can affect the
diversity-stability relationship, we simulated the model with four
species under two scenarios for each coexistence mechanism. First, we
allowed the variance of the environment to determine how many species
can coexist, akin to a community assembly experiment with a species pool
of four species. This required simulating communities with all species
initially present across a gradient of annual resource variability (for
relative nonlinearity) or environmental cue variability (for the storage
effect). Second, we chose parameter values that allowed coexistence of
all four species and performed species removals. The two simulation
experiments correspond to (i) sampling ecosystem function across a
natural gradient of species richness and (ii) sampling ecosystem
function across diversity treatments within a site.

To understand how increasing environmental variability will impact
ecosystem stability when coexistence is fluctuation-dependent, we
simulated the model over a range of environmental cue variability and
species pool sizes. Thus, for each size of species pool (1, 2, 3, or 4
species), we simulated the model at 15 evenly-spaced levels of
environmental cue (range = 0.1,2) for the storage effect and 25
evenly-spaced levels of resource variability (range = 0.1,1.4) for
relative nonlinearity. We also explored the influence of species
asymmetries in competitive ability and species' correlations of
environmental responses in the storage effect model. Under the storage
effect, if all species are perfectly symmetrical, that is, there is no
superior competitor, then coexistence is fluctuation independent. We use
one such parameterization of our model to contrast the response of
ecosystem stability in fluctuation-dependent and fluctation-independent
communities to environmental variation. Likewise, under relative
nonlinearity, species' resource response curves (Fig. SX) reflect traits
that determine the intrinsic stability of each species. Therefore, we
ran two sets of simulations for relative nonlinearity: one where the
species pools increased from stable to unstable species and vice versa.
For example, if species A is the most stable species and species D is
the least stable we ran simulations where the species pool increased
from one to four species as A then B then C then D. We then ran
simulations with that order reversed.

All simulations were run for 5,000 seasons with 20-day growing seasons.
We averaged biomass over the growing season, and those yearly values
were used to calculate total community biomass in each year. After
discarding an initial 500 seasons to reduce transient effects on our
results, we calculated the coffecient of variation (\emph{CV}) of summed
species biomass through time. Therefore, in our results we refer to
ecosystem variability, which is the inverse of ecosystem stability. We
calculated species richness as the number of species whose average
biomass was greater than 1 over the course of the simulation. Parameter
values for specific results are given in figure captions. Within-season
dynamics were solved given initial conditions using the package
\texttt{deSolve} (Soetaert et al. 2010) in \texttt{R} (Team 2013). All
model code has been deposited on Figshare (\emph{link}) and is available
on GitHub at \url{http://github.com/atredennick/Coexistence-Stability}.

\subsection{RESULTS AND DISCUSSION}\label{results-and-discussion}

@@

The direction of the diversity-variability relationship can be positive
and negative when species coexistence is maintained by
fluctuation-dependent mechanisms (Fig. 2). Ecosystem variability is
positively correlated with species richness when species richness is
measured across a gradient of environmental variability, which maintains
diversity and promotes ecosystem variability (Fig. 2A,C). If
environmental conditions are sufficient to maintain coexistence,
removing species increases ecosystem variability (Fig. 2B,D). Thus, our
results both confirm and contrast with theoretical and experimental
findings that diversity begets stability.

When we held environmental variability constant and removed species, we
produced the typical negative diversity-variability relationship (Fig.
2B,D), consistent with theoretical expecations from models with species
coexistence maintained by fluctuation-dependent mechanisms. Likewise,
our results from the species removal simulations are consistent with
results from biodiversity-ecosystem functioning experiments showing a
negative relationship between species richness and ecosystem
variability. This is encouraging because species almost certainly
coexist by some combination of fluctuation-independent (e.g., resource
partitioning) and fluctuation-dependent mechanisms. By extending theory
to communities where species richness is explicitly maintained by
temporal variability, we have gained confidence that experimental
findings are generalizable to many communities. In other words, in local
settings where environmental variability is relatively homogenous,
reductions in the number of species will reduce the stability of
ecosystem functioning, regardless of how coexistence is maintained.

When we allowed a gradient of environmental variability to determine
species coexistence, we discovered a positive relationship between
species richness and ecosystem variability (Fig. 2A,C). While surprising
when viewed through the lens of previous theory and experimental
findings, such a relationship is a direct consequence of how diversity
can be maintained in fluctuating environments. The storage effect and
relative nonlinearity both require envrionmental fluctuations to allow
niche differentiation between species pairs (Chesson 2000). Therefore,
species coexistence gains strength, for both mechanisms, as the
environment becomes more variable (Fig. SX).

@@

Species coexistence in real ecological communities probably emerges from
some combination of fluctuation-independent and fluctuation-dependent
mechanisms (Chesson 2000, Clark et al. 2010). Likewise, environmental
conditions in real ecosystems are unlikely to change only in their
variability without an associated change in the mean (Avolio et al.
2015). Therefore, environmental change has the potential to alter the
niche and fitness differences among species in multiple ways, some of
which were not present in our current analysis. Mean changes in
environmental conditions could reorder competitive hierarchies
(Klanderud and Totland 2005) and/or alter the availability of niches
(Harpole et al. 2016). Associated changes in ecosystem stability will
depend upon the magnitude of environmental change, each species'
response to the particular environmental driver, and biotic interactions
(Hallett et al. 2014). Thus, it is becoming clear that understanding how
ecosystem stability will respond to global change will require a
trait-based approach.

@@

Adler, P. B., and J. M. Drake. 2008. Environmental variation, stochastic
extinction, and competitive coexistence. The American Naturalist
172:186--195.

@@

Chesson, P. 2000. Mechanisms of Maintenance of Species Diversity. Annual
Review of Ecology and Systematics 31:343--366.

>>>>

MacArthur (1955), Elton (1958), and even Darwin (Turnbull et al. 2013)
recognized that species can compensate for each other and stabilize
ecosystem functioning in fluctuating environments. This idea underlies
the ``insurance hypothesis'' (Yachi and Loreau 1999), which suggests
stability increases with diversity because species respond dissimilarly
to environmental conditions, broadening the range of conditions under
which the community maintains function (Loreau 2010). Diverse models all
predict a positive relationship between species richness and ecosystem
stability (Lehman and Tilman 2000, Ives and Hughes 2002, Loreau and {{de
Mazancourt}} 2013), and experimental tests tend to support such a
prediction (Tilman et al. 2006, Hector et al. 2010).

However, the ability of biodiversity-ecosystem functioning (BEF)
experiments to represent real-world dynamics is the subject of debate
(Wardle 2016, Eisenhauer et al. 2016). Much of the debate centers around
the fact that species losses in BEF experiments are not allowed to be
offset by species gains. Theoretical work on diveristy-stability
relationships typically suffers from the same limitation, ignoring the
processes by which temporal environmental variability can promote
coexistence and increase species richness (Loreau 2010, but see Chesson
et al. 2001).

Fluctuating environmental conditions are an important ingredient for
stable species coexistence, both in theoretical models (Chesson 2000a,
Chesson et al. 2004) and in natural communities (C{á}ceres 1997,
Descamps-Julien and Gonzalez 2005, Adler et al. 2006, Angert et al.
2009). Such ``fluctuation-dependent'' coexistence requires that species
have unique environmental responses and that environmental conditions
vary so that each species in a community experiences favorable and
unfavorable conditions, which prevents competitive exclusion (Chesson
2000a). When coexistence is maintained by a fluctuation-dependent
mechanism, an increase in environmental variability might lead to an
increase in species richness and, consequently, an increase in ecosystem
stability. However, increasing environmental variability may also
decrease ecosystem stability by increasing the fluctuations of
individual species, regardless of species richness.

The countervailing effects of environmental variability present an
interesting paradox: increasing variability should decrease ecosystem
stability, but the potential for an increase in richness might offset
the decrease in stability. Such a paradox complicates predictions about
how ecosystems will respond as environmental conditions exceed
historical ranges of variability. The unknown net effect of environment
variability may be reflected in the mixed results from empirical studies
on the diversity-stability relationship. Observational tests of the
diversity-stability relationship, which require sampling across natural
diversity gradients, have yielded positive (Hautier et al. 2014),
neutral (Valone and Hoffman 2003, Cusson et al. 2015), and negative
(Sasaki and Lauenroth 2011) relationships. In a meta-analysis of
diversity-stability relationships based on observational studies in
terrestrial ecosystems, Jiang and Pu (2009) found no significant
evidence for an effect of species richness on ecosystem stability. The
idiosyncratic results of these empirical studies contrasts with the
consistent conclusions from theoretical work.

The gap between theoretical expectations and empirical results of
richness-stability relationships reflects the divergence of theory
developed to explain species coexistence and theory developed to explain
diversity and stability. One reason these two disciplines have diverged
is because they have focused on slightly different questions.
Diversity-stability studies typically ask how ecosystem variability
responds to different levels of species richness at a given level of
environmental variability (reviewed in Kinzig et al. 2001, Loreau 2010),
whereas coexistence studies ask how the long term stability of species
coexistence responds to different levels of environmental variability
(Chesson and Warner 1981).

To reconcile these two perspectives, we extend theory on the
relationship between species richness and ecosystem stability to cases
in which species coexistence explicitly depends on environmental
fluctuations and species-specific responses to environmental conditions.
We focus on the temporal storage effect and relative nonlinearity using
a general consumer-resource model. We use the model to investigate three
questions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Does the direction of the diversity-stability relationship remain
  positive when species coexistence is fluctuation-dependent?
\item
  When species coexistence is fluctuation-dependent, how does increasing
  environmental variability impact ecosystem stability?
\item
  Do our answers to the previous two questions depend on the specific
  fluctuation-dependent coexistence mechanism (i.e., storage effect
  vs.~relative nonlinearity)?
\end{enumerate}

@@

For the storage effect to operate, we need species-specific responses to
environmental variability, positive environment-competition (EC)
covariance, and subadditive population growth (Chesson 1994, 2000b). EC
covariance limits population growth of residents: in a favorable year,
they are limited by strong intraspecific competition because their
populations are large. However, a species at very low abundance can
escape the limitations of EC covariance if it experiences favorable
conditions when the residents it is competiting against experience
(relatively) unfavorable conditions. Subadditive population growth
buffers populations against large population decreases in unfavorable
years.

Our model includes subadditive population growth because we include a
dormant stage with very low death rates, which limits large population
declines. Our model includes EC covariance through the effect of the
dormant-to-live transition rate, \(\gamma\), on population growth. A
high \(\gamma\) year for a species represents a good \emph{E} year in
which live biomass, and thus resource uptake, is above average. These
conditions correspond to \emph{C} also being above average because a
common species, by definition, cannot escape intraspecific competition.

In our model, rare species are rare because they are inferior
competitors. So, given adequate variability in \(\gamma\), the inferior
competitor can persist because its rarity allows it to take advantage of
favorable envrionmental conditions. We created competitive hierarchies
in the storage effect version of the model by altering species' biomass
conversion efficiencies (\(\epsilon\))

@@

\textcolor{red}{YOU NEED TO REMIND READERS ABOUT THE STANDARD RESOURCE COMPETITION MODELS, MACARTHUR, TILMAN, ETC. IN GENERAL, I THINK THIS PARAGRAPH NEEDS TO BE EXPANDED AND EXPLAINED MORE SLOWLY, STEP BY STEP.}

@@

To explore how fluctuation-dependent coexistence can affect the
diversity-stability relationship, we simulated the model with four
species under two scenarios for each coexistence mechanism. First, we
allowed the variance of the environment to determine how many species
can coexist, akin to a community assembly experiment with a species pool
of four species. To do this, we simulated communities with all species
initially present across a gradient of annual resource variability (for
relative nonlinearity) or environmental cue variability (for the storage
effect). Second, we chose parameter values that allowed coexistence of
all four species and then performed species removals, akin to a
biodiversity-ecosystem function experiment. The two simulation
experiments correspond to (i) sampling ecosystem function across a
natural gradient of species richness and (ii) sampling ecosystem
function across diversity treatments within a site.

To understand how increasing environmental variability will impact
ecosystem stability when coexistence is fluctuation-dependent, we
simulated the model over a range of species pool sizes and environmental
cue or resource variability. For each size of species pool (1, 2, 3, or
4 species), we simulated the model at 15 evenly-spaced levels of
environmental cue (range = 0.1,2) for the storage effect and 25
evenly-spaced levels of resource variability (range = 0.1,1.4) for
relative nonlinearity. We also explored the influence of asymmetries in
species' competitive abilities and correlations in species'
environmental responses within the storage effect model.

Under relative nonlinearity, species' resource response curves (Fig. SX)
reflect traits that determine the intrinsic stability of each species.
Therefore, we ran two sets of simulations for relative nonlinearity: one
where the species pools increased from stable to unstable species and
vice versa. For example, if species A is the most stable species and
species D is the least stable we ran simulations where the species pool
increased from one to four species as A then B then C then D. We then
ran simulations with that order reversed.

All simulations were run for 5,000 seasons with 20-day growing seasons.
We averaged biomass over the growing season, and those yearly values
were used to calculate total community biomass in each year. After
discarding an initial 500 seasons to reduce transient effects on our
results, we calculated the coffecient of variation (\emph{CV}) of summed
species biomass through time, which represents ecosystem variability,
the inverse of ecosystem stability. We calculated species richness as
the number of species whose average biomass was greater than 1 over the
course of the simulation. Parameter values for specific results are
given in figure captions. Within-season dynamics were solved given
initial conditions using the package \texttt{deSolve} (Soetaert et al.
2010) in \texttt{R} (Team 2013). All model code has been deposited on
Figshare (\emph{link}) and is available on GitHub at
\url{http://github.com/atredennick/Coexistence-Stability}.

\subsection{RESULTS}\label{results}

Our simulation experiments show that the direction of the
diversity-variability relationship can be positive or negative when
species coexistence is maintained by fluctuation-dependent mechanisms
(Fig. 2). Ecosystem variability is positively correlated with species
richness when species richness is measured across a gradient of
environmental variability. The variability promotes diversity (Fig. SX)
but also leads to increased ecosystem variability (Fig. 2A,C). On the
other hand, holding environmental conditions constant (but at levels
sufficient for coexistence), removing species increases ecosystem
variability (Fig. 2B,D).

\subsection{DISCUSSION}\label{discussion}

@@

Theory developed for biodiversity-ecosystem function experiments
emphasizes that increases in species richness should reduce ecosystem
variability. , consistent with theoretical expectations from models in
which species coexistence maintained by fluctuation-dependent
mechanisms, and consistent with Species almost certainly coexist by some
combination of fluctuation-independent (e.g., resource partitioning) and
fluctuation-dependent mechanisms. By extending theory to communities
where species richness is explicitly maintained by temporal variability,
we have gained confidence that experimental findings are generalizable
to many communities. In other words, in local settings where
environmental variability is relatively homogenous, reductions in the
number of species will reduce the stability of ecosystem functioning,
regardless of how coexistence is maintained.

@@

Species coexistence in real ecological communities probably emerges from
some combination of fluctuation-independent and fluctuation-dependent
mechanisms (Chesson 2000a, Clark et al. 2010). Likewise, environmental
conditions in real ecosystems are unlikely to change only in their
variability without an associated change in the mean (Avolio et al.
2015). Therefore, environmental change has the potential to alter the
niche and fitness differences among species in multiple ways, some of
which were not present in our current analysis. Mean changes in
environmental conditions could reorder competitive hierarchies
(Klanderud and Totland 2005) and/or alter the availability of niches
(Harpole et al. 2016). Associated changes in ecosystem stability will
depend upon the magnitude of environmental change, each species'
response to the particular environmental driver, and biotic interactions
(Hallett et al. 2014). Thus, it is becoming clear that understanding how
ecosystem stability will respond to global change will require a
trait-based approach.

@@

@@

Chesson, P. 2000a. Mechanisms of Maintenance of Species Diversity.
Annual Review of Ecology and Systematics 31:343--366.

Chesson, P. 2000b. General theory of competitive coexistence in
spatially-varying environments. Theoretical population biology
58:211--37.

Chesson, P. L. 1994. Multispecies Competition in Variable Environments.
Theoretical Population Biology 45:227.
